---
title: "code"
author: "DBPR"
date: "11/26/2023"
output: html_document
---
This code details all the analyses and results of the study "Marine fishes experiencing high-velocity range shifts are not climate change ‘winners’"
For more information please contact DBPR


#Packages
```{r}
library(tidyverse)
library(glmmTMB)
```

#Loading the data
```{r}
data=read.csv("data.csv")
```

Use decimal numbers
```{r}
options(scipen=999)
```

#main effect
The relationship between population trends and range shift velocity.
Population trends (slopes) based on linear models with scaled abundance.
```{r}
#model
test_main_effect=glmmTMB(data=data,
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_effect)
MuMIn::r.squaredGLMM(test_main_effect)
#Model diagnostics
sjPlot::plot_model(test_main_effect,type="diag")
#AIC
AIC(test_main_effect)
```

##mean shift velcoity
Test the main effect using the mean shift velocity estimate
```{r}
#model
test_main_effect_mean_shift=glmmTMB(data=data,
formula=slope~mean_shift+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_effect_mean_shift)
MuMIn::r.squaredGLMM(test_main_effect_mean_shift)
#Model diagnostics
sjPlot::plot_model(test_main_effect_mean_shift,type="diag")
#AIC
AIC(test_main_effect_mean_shift)
```

##remove extreme range shift velocities
```{r}
#shift of minimum sm
quantile(data$shift_of_min_sm)[[4]]
IQR=quantile(data$shift_of_min_sm)[[4]]-quantile(data$shift_of_min_sm)[[2]]
min=quantile(data$shift_of_min_sm)[[2]]-(IQR*1.5)
max=quantile(data$shift_of_min_sm)[[4]]+(IQR*1.5)
#model
test_main_effect_rm_ex_sh=glmmTMB::glmmTMB(data=data %>% filter(shift_of_min_sm<=max & shift_of_min_sm>=min),
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n) 
#Model Summary
summary(test_main_effect_rm_ex_sh)
MuMIn::r.squaredGLMM(test_main_effect_rm_ex_sh)
#Model diagnostics
sjPlot::plot_model(test_main_effect_rm_ex_sh,type="diag")
```

##Leading edge range shift velocities
Using only leading edge estimates from BioShifts
```{r}
data_leading=read.csv("data_leading.csv")
#model
test_main_effect_leading=glmmTMB(data=data_leading,
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_effect_leading)
MuMIn::r.squaredGLMM(test_main_effect_leading)
#Model diagnostics
sjPlot::plot_model(test_main_effect_leading,type="diag")
```

##Trailing edge range shift velocities
Using only trailing edge estimates from BioShifts
```{r}
data_trailing=read.csv("data_trailing.csv")
#model
test_main_effect_trailing=glmmTMB(data=data_trailing,
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_effect_trailing)
MuMIn::r.squaredGLMM(test_main_effect_trailing)
#Model diagnostics
sjPlot::plot_model(test_main_effect_trailing,type="diag")
```

##species level
```{r}
#Create species level shift with stat. summary
species_level_shifts=data %>% 
  group_by(species) %>% 
  summarise(mean_shift_of_min_sm=mean(shift_of_min_sm),
            n=n())
#Create species level data
data_sp_level=data %>% 
  group_by(species,family) %>% 
  summarise(n_es=n(),
            mean_es=mean(slope))  %>%
  inner_join(species_level_shifts,by="species") %>% 
  ungroup() %>% 
  mutate(across(c(4:5), round, 3),
         sqrt_n_es=sqrt(n_es))
#Model
test_sp_level=glmmTMB::glmmTMB(data=data_sp_level,
formula=mean_es~mean_shift_of_min_sm+
  (1|family),
family = 'gaussian',
weights =sqrt_n_es)
#Model Summary
summary(test_sp_level)
MuMIn::r.squaredGLMM(test_sp_level)
#Model diagnostics
sjPlot::plot_model(test_sp_level,type="diag")
```

##Replace weights
Replace weights and use the square root transformed inverse SE of the slope
```{r}
#model
test_main_replace_w=glmmTMB::glmmTMB(
  data=data %>% 
    mutate(sqrt_inverse_se=sqrt((1/(se_slope+0.0003)))),
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_inverse_se)
#Model Summary
summary(test_main_replace_w)
MuMIn::r.squaredGLMM(test_main_replace_w)
#Model diagnostics
sjPlot::plot_model(test_main_replace_w,type="diag")
```

##The effect of estimating range shifts from occurrence or abundance
```{r}
#model
test_main_abundance_based=glmmTMB::glmmTMB(
  data=data,#use BioShifts estimated from abundance data
formula=slope~shift_of_min_sm*BioShifts_prab+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_abundance_based)
MuMIn::r.squaredGLMM(test_main_abundance_based)
#Model diagnostics
sjPlot::plot_model(test_main_abundance_based,type="diag")
```

##Controling for Bioshifts data quality levels
###Resurveyd
```{r}
#model
test_res=glmmTMB::glmmTMB(
  data=data %>% 
    filter(BioShifts_quality %in%"RESURVEYED"),
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_res)
MuMIn::r.squaredGLMM(test_res)
#Model diagnostics
sjPlot::plot_model(test_res,type="diag")
```

###Actual data (i.e., 'LOW')
```{r}
#model
test_acdat=glmmTMB::glmmTMB(
  data=data%>% 
    filter(BioShifts_quality %in%"LOW"),
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_acdat)
MuMIn::r.squaredGLMM(test_acdat)
#Model diagnostics
sjPlot::plot_model(test_acdat,type="diag")
```

##Ocean regions
Categorize by oceans
```{r}
data=data %>% 
  mutate(ocean=case_when(latitude_mean<=90 & 
                         latitude_mean>=0 &
                         longitude_mean<=20 &
                         longitude_mean>=-35~"NE. Atlantic",
                         latitude_mean<=90 & 
                         latitude_mean>=0 &
                         longitude_mean<(-35) &
                         longitude_mean>=-100~"NW. Atlantic",
                         latitude_mean<=90 & 
                         latitude_mean>=20 &
                         longitude_mean<100 &
                         longitude_mean>=-175~"NE. Pacific",
                         latitude_mean<=3 & 
                         latitude_mean>=-48 &
                         longitude_mean<=180 &
                         longitude_mean>=110~"Oceania",
                         TRUE~"Other"))
```

Test
```{r}
#model
test_main_ocean=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*ocean+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_ocean)
MuMIn::r.squaredGLMM(test_main_ocean)
#Model diagnostics
sjPlot::plot_model(test_main_ocean,type="diag")
performance::check_collinearity(test_main_ocean)
#Table with model results
sjPlot::tab_model(test_main_ocean,digits = 3)
```

##Test with R^2 >= 0.5
```{r}
#Model
test_main_r2=glmmTMB::glmmTMB(data=data %>%
                                filter(r.squared>= .5),
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_r2)
MuMIn::r.squaredGLMM(test_main_r2)
#Model diagnostics
sjPlot::plot_model(test_main_r2,type="diag")
```

##Test percent change in abundance
Population trends are based on GLMs with poisson error distribution.
```{r}
#Model
test_percent_change=glmmTMB::glmmTMB(
  data=data %>%
    filter(ABUNDANCE_TYPE%in%"Count"),
formula=ln_ratio_fitted~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n) 
#Model Summary
summary(test_percent_change)
MuMIn::r.squaredGLMM(test_percent_change)
#Model diagnostics
sjPlot::plot_model(test_percent_change,type="diag")

##ggeffects
#predict table for back transformed values
test_percent_change_gg=ggeffects::ggpredict(test_percent_change,                                           terms="shift_of_min_sm [-21:43 by=1]") %>% 
  as_tibble() %>% 
  mutate(exp_predicted=exp(predicted) %>% 
           round(digits=4),
         exp_conf.high=exp(conf.high)%>% 
           round(digits=4),
         exp_conf.low=exp(conf.low)%>% 
           round(digits=4),
         exp_prediceted_percent=(exp_predicted-1)*100,
         exp_conf.low_percent=(exp_conf.low-1)*100,
         exp_conf.high_percent=(exp_conf.high-1)*100) %>% 
  select(x,exp_prediceted_percent,exp_conf.low_percent,exp_conf.high_percent)
```

###Example species
Example species of extreme range shift velocity and their associated change in population size
```{r}
data %>% 
  group_by(species) %>% 
  filter(shift_of_min_sm>=17) %>% 
  summarise(populations=n_distinct(unique_id),
            shift=min(shift_of_min_sm),
            mean_decline=(100*(1-exp(mean(ln_ratio_fitted)))) %>% 
              round(digits=2)) %>% 
  arrange(mean_decline,shift,populations)
#Positive 'mean_decline' values denote negative percentage change in population size
```

#Poleward vs. equatorward range shift velocities
##Poleward
```{r}
#model
test_main_effect_poleward=glmmTMB(
  data=data %>% 
    filter(shift_of_min_sm>0),
formula=slope~shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_effect_poleward)
MuMIn::r.squaredGLMM(test_main_effect_poleward)
#Model diagnostics
sjPlot::plot_model(test_main_effect_poleward,type="diag")
```

##Equatorward
Here we will use absolute shift values for interpretation purposes.
```{r}
#model
test_main_effect_equatorward=glmmTMB(
  data=data %>% 
    filter(shift_of_min_sm<0) %>% 
    mutate(abs_shift=abs(shift_of_min_sm)),
formula=slope~abs_shift+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_main_effect_equatorward)
MuMIn::r.squaredGLMM(test_main_effect_equatorward)
#Model diagnostics
sjPlot::plot_model(test_main_effect_equatorward,type="diag")
```

#Spatial and thermal position
##The effect of spatial position
```{r}
#model
test_spat_pos=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*spatial_pos+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_spat_pos)
MuMIn::r.squaredGLMM(test_spat_pos)
#Model diagnostics
sjPlot::plot_model(test_spat_pos,type="diag")
performance::check_collinearity(test_spat_pos)
#AIC
AIC(test_spat_pos)
```

###Test hump-shaped pattern
In an extreme shift scenario, former poleward populations might become the new equatorward populations.
```{r}
#model
test_non_linear=glmmTMB::glmmTMB(
  data=data %>%
            filter(spatial_pos>0,
                  shift_of_min_sm>0),#due to the quadratic transformation we eliminate negative shift values that cannot be back-transformed.
formula=slope~I(shift_of_min_sm^2)+shift_of_min_sm+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_non_linear)
MuMIn::r.squaredGLMM(test_non_linear)
#Model diagnostics
performance::check_collinearity(test_non_linear)
sjPlot::plot_model(test_non_linear,type="diag")
```

##the effect of thermal position
```{r}
#model
test_therm_pos=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*thermal_pos+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_therm_pos)
MuMIn::r.squaredGLMM(test_therm_pos)
#Model diagnostics
sjPlot::plot_model(test_therm_pos,type="diag")
performance::check_collinearity(test_therm_pos)
```

###Subsetting by species habitat (i.e., benthic, benthopelagic, and pelagic)
####Benthic
```{r}
#model
test_therm_pos_b=glmmTMB::glmmTMB(
  data=data %>% 
  filter(habitat%in%c("b")),
formula=slope~shift_of_min_sm*thermal_pos+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_therm_pos_b)
MuMIn::r.squaredGLMM(test_therm_pos_b)
#Model diagnostics
sjPlot::plot_model(test_therm_pos_b,type="diag")
performance::check_collinearity(test_therm_pos_b)
```

####Benthopelagic + benthic
```{r}
#model
test_therm_pos_bp=glmmTMB::glmmTMB(
  data=data %>% 
  filter(habitat%in%c("bp","b")),
formula=slope~shift_of_min_sm*thermal_pos+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_therm_pos_bp)
MuMIn::r.squaredGLMM(test_therm_pos_bp)
#Model diagnostics
sjPlot::plot_model(test_therm_pos_bp,type="diag")
performance::check_collinearity(test_therm_pos_bp)
```

####Pelagic
```{r}
#model
test_therm_pos_p=glmmTMB::glmmTMB(
  data=data %>% 
  filter(habitat%in%c("p")),
formula=slope~shift_of_min_sm*thermal_pos+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_therm_pos_p)
MuMIn::r.squaredGLMM(test_therm_pos_p)
#Model diagnostics
sjPlot::plot_model(test_therm_pos_p,type="diag")
performance::check_collinearity(test_therm_pos_p)
```

#latitude, fishing, and spatiotemporal mismatches
##Pop. trends against abs. latitude
```{r}
#model
test_abs_lat=glmmTMB::glmmTMB(data=data,
formula=slope~abs_latitude_mean+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_abs_lat)
MuMIn::r.squaredGLMM(test_abs_lat)
#Model diagnostics
sjPlot::plot_model(test_abs_lat,type="diag")
AIC(test_abs_lat)
```

##Pop. trends against abs. lat. * spat. pos
```{r}
#model
test_abs_lat_sp=glmmTMB::glmmTMB(data=data,
formula=slope~abs_latitude_mean*spatial_pos+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_abs_lat_sp)
MuMIn::r.squaredGLMM(test_abs_lat_sp)
#Model diagnostics
sjPlot::plot_model(test_abs_lat_sp,type="diag")
#AIC
AIC(test_abs_lat_sp)
```

##pop. trends against vulnerability*shift velocity
```{r}
#model
#Two-way interaction term
test_vul_shift=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*scale(Vulnerability)+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_vul_shift)
MuMIn::r.squaredGLMM(test_vul_shift)
#Model diagnostics
sjPlot::plot_model(test_vul_shift,type="diag")
performance::check_collinearity(test_vul_shift)
```

##pop. trends against vulnerability*shift velocity*spatial pos.
```{r}
#model
#Two-way interaction term
test_vul_sp_shift=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*scale(Vulnerability)*spatial_pos+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_vul_sp_shift)
MuMIn::r.squaredGLMM(test_vul_sp_shift)
#Model diagnostics
sjPlot::plot_model(test_vul_sp_shift,type="diag")
performance::check_collinearity(test_vul_sp_shift)
```

##Pop. trends against commercial level
```{r}
#model
test_pop_com=glmmTMB::glmmTMB(data=data,
formula=slope~commercial_two_level+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_pop_com)
MuMIn::r.squaredGLMM(test_pop_com)
#Model diagnostics
sjPlot::plot_model(test_pop_com,type="diag")
```

##Pop. trends against commercial level
```{r}
#model
test_pop_com_shift=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*spatial_pos*commercial_two_level+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_pop_com_shift)
MuMIn::r.squaredGLMM(test_pop_com_shift)
#Model diagnostics
performance::check_collinearity(test_pop_com_shift)
sjPlot::plot_model(test_pop_com_shift,type="diag")
```

#Spatial and temporal mismatch
##spatial mismatch
The spatial distance is the distance (in kilometer) between the location of a BioTIME population and the location in which the range shift velocity centroid is.
```{r}
#model
test_spat_mis=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*scale(spatial_pos)*scale(bt_bs_dist_of_min_sm)+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_spat_mis)
MuMIn::r.squaredGLMM(test_spat_mis)
#Model diagnostics
sjPlot::plot_model(test_spat_mis,type="diag")
performance::check_collinearity(test_spat_mis)
```

##Temporal mismatch
The temporal distance is the distance (in years) between the year in which a BioTIME population assessment has begun and the year in which the range shift velocity assessment has begun.
```{r}
#model
test_tm=glmmTMB::glmmTMB(data=data,
formula=slope~shift_of_min_sm*spatial_pos*scale(temporal_mismatch)+
  (1|family)+
  (1|study_id),
family = 'gaussian',
weights =sqrt_n)
#Model Summary
summary(test_tm)
MuMIn::r.squaredGLMM(test_tm)
#Model diagnostics
sjPlot::plot_model(test_tm,type="diag")
performance::check_collinearity(test_tm)
```